---
title: "Introduction"
author: "Doug Antibus"
date: "2023-12-28"
output: 
  html_document:
    toc: true
    number_sections: false
    theme:
      version: 5
      bootswatch: pulse
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
source("00_libraries.r")
```

### Introduction to case study

This is a case study which I completed for the [*Google Data Analytics Professional
Certificate*](https://www.coursera.org/professional-certificates/google-data-analytics). The goal
of this case study is to assist **Cyclistic**, a fictional bike-share `r icon_style(fontawesome("bicycle", style = "solid"), scale = 2, fill = "purple")` company offering rentals in Chicago, IL.  

Cyclistic recognizes two types of riders:

- **Members** are users who hold annual memberships

- **Casual riders** are users who use single-ride passes or day passes

Cyclistic finds that members are more profitable than casual riders, hence they would
like to create marketing strategies to convert casual riders to members.  To do so,
they want to understand differences in how members and casual riders use the bike-share 
service.  

In this study, I analyzed ridership data spanning from November 2022 - October 2023
to identify differences in member and casual ridership.  I wanted to include a 
complete year of data in my analysis in order to uncover seasonal differences that
might exist between the two types of riders.  The report for this case study will
cover several phases of data analysis:

- Downloading data

- Inspecting, cleaning, and data transformation

- Exploratory data analysis

- Plotting

- Modeling to find variables that predict member vs. casual rider type


### Downloading data

Raw data files were downloaded from here: https://divvy-tripdata.s3.amazonaws.com/index.html

Each .zip file represents a month of trip data.  I downloaded files spanning the 
time period from November 2022 - October 2023.  

The below code sets the url root for trip data files.  It then loads necessary
libraries.  Finally, we run a script to download files, placing them in the "data" subfolder 
of our project and then unzipping them.  


```{r url}
url_root <- "https://divvy-tripdata.s3.amazonaws.com/"
```

```{r libraries, message = FALSE, warning = FALSE}
source(here("00_libraries.r"))
```

```{r, message = FALSE, eval = FALSE, cache=TRUE}
source(here("00_file_downloads.r"))
```

### Combining data

For this dataset, each month is represented by a single file.  We want to combine
all the monthly data into a single data frame.  The below code sources the 
script "01_reading_combining.r".  This script reads each **.csv** file into a list
then uses `dplyr::bind_rows` to bind them into a single data frame.  The resulting 
data frame has approximately **5.6 million** records.


```{r combine, eval = FALSE, cache=TRUE}
source(here("01_reading_combining.r"))
```

We can inspect the raw data. It contains the following variables:

- the id column `ride_id`

- `rideable_type` which represents the bike type (electric, classic, or docked)

- trip start and end datetimes

- start and end station names and station id's

- start and end longitude and latitude

- `member_casual`, the outcome variable of interest, whether the rider is a member or 
casual rider

```{r, cache=TRUE}

if (file_exists(here("data", "combined_data_raw.csv"))) {
  combined_data_raw <- read_csv(here("data", "combined_data_raw.csv"))
}

if (exists("combined_data_raw")) glimpse(combined_data_raw)
```

### Session information

```{r}
sessionInfo()
```
