---
title: "Exploratory Data Analysis"
author: "Doug Antibus"
date: "2023-12-28"
output: 
  html_document:
    toc: true
    number_sections: false
    theme:
      version: 5
      bootswatch: pulse
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
source("00_libraries.r")
```

### Exploratory Data Analysis

With the data cleaned and transformed, we can start asking basic questions in preparation
for plotting and modeling.  We start by reading in the most recent data file `bikes_2.csv` 
and omitting rows containing NA values.

```{r, cache=TRUE}
bikes <- read_csv(here("data", "bikes_2.csv")) %>% na.omit()
glimpse(bikes)
```

#### Overall ridership

First, we examine the overall distribution of member and casual trips. Member trips 
outnumber casual trips 64% to 36%.  It will be important to consider this difference
for subsequent analysis.

```{r}
mc <- bikes %>% group_by(member_casual) %>% 
  summarize(n = n()) %>% 
  mutate(percentage = round(100 * n / sum(n), 2))
mc
```
### Bike types

There are 3 `rideable_type`s (types of bikes) in the dataset: classic, electric, and docked.  The table below summarizes the abundance of trips for each bike type in the entire dataset.  Docked bikes represent a small proportion (2%) of trips.  Classic bike trips outnumber electric bike trips by 61% to 37%.

```{r}
ride_type <- bikes %>% group_by(rideable_type) %>% 
  summarize(n = n()) %>% 
  mutate(percentage = round(100 * n / sum(n), 2))
ride_type
```

We break down trips by both bike type and member/casual (6 possible combinations). For each combination we calculate the percentage of total trips, as well as the percentage of trips for that membership type.  We find the following:

- only casual riders use docked bikes

- members prefer classic bikes: trips by members are enriched for classic rides compared to the dataset as a whole (64% vs. 61%; see preceeding table)

- conversely, casuals prefer electric bikes: trips by casuals are *less* enriched 
for classic bikes than the dataset as a whole


```{r}
mc_ride_type <- bikes %>% group_by(rideable_type, member_casual) %>% 
  summarize(n = n()) %>% 
  ungroup() %>% 
  mutate(pct_of_total = round(100 * n / sum(n), 2)) %>% 
  arrange(member_casual) %>% 
  ungroup() %>% 
  group_by(member_casual) %>% 
  mutate(pct_by_membership_type = round(100 * n / sum(n), 2))
mc_ride_type
```
### Who takes more round trips?

The overall occurrence of round trips is fairly low.  However, casual riders are 
over twice as likely as members to take round trips.  9.7% of casual trips are round trips vs. 4.6% of member trips.  

One hypothesis we can consider is that members largely represent commuters (who 
find value in a membership as they ride frequently), while casuals represent tourists.  This observation is consistent with that hypothesis: commuters might largely want to go from "Point A to Point B" so they would be expected to take 
round trips less frequently than tourists, who might want to return to a point of 
departure.

```{r}
round_trips <- bikes %>% 
  group_by(member_casual) %>% 
  summarize(total_trips = n(), 
            round_trips = sum(round_trip == "yes"),
            pct_round_trips = round(100 * round_trips / total_trips, 2))
round_trips
```

### Who takes more trips on holidays?

Holiday trips are fairly rare.  However, casual riders seem to take more trips on holidays (3.6% of casualm trips are on holidays vs. 2.8% of member trips).  A chi-square test shows that this difference is significant with a p-value of 0.

```{r}
holiday <- bikes %>% 
  group_by(member_casual, is_holiday) %>% 
  summarize(n = n()) %>% 
  group_by(member_casual) %>% 
  mutate(pct_of_rider_type = round(100 * n / sum(n), 2))
holiday
```

```{r}
chisq_test(bikes, is_holiday ~ member_casual)
```

Overall conclusions from the EDA include the following:

- members take more trips than casual riders

- members prefer classic bikes while casual riders prefer electric bikes

- casual riders are more likely to take round trips than members (although
not many round trips are taken overall)

- casual riders are more likely to take trips on holidays (although holiday trips
are not abundant overall)


