---
title: "Station latitude and longitude"
author: "Doug Antibus"
date: "2023-12-28"
output: 
  html_document:
    toc: true
    number_sections: false
    theme:
      version: 5
      bootswatch: pulse
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(cache = TRUE)
source("00_libraries.r")
```

### Transforming geolocation data 

For future analyses (e.g. plotting lat and lon data), it will be helpful
to summarize mean latitude and longitude by station.  We start by 
reading in the most recent data file `bikes_1.csv`

```{r import, cache = TRUE}
bikes <- read_csv(here("data", "bikes_1.csv"))
```

Next, pull out data for trip starts and trip ends separately, remove NA values
then bind these into a single data frame called `stations`

```{r, cache = TRUE}
starts <- bikes %>% select(ride_id, start_name, start_id, start_lat, start_lon) %>% 
  rename_with(function(x) str_remove(x, "start_")) %>%
  mutate(type = "start") %>% 
  na.omit()

ends <- bikes %>% select(ride_id, end_name, end_id, end_lat, end_lon) %>% 
  rename_with(function(x) str_remove(x, "end_")) %>% 
  mutate(type = "end") %>% 
  na.omit() 

stations <- bind_rows(starts, ends)
glimpse(stations)
```

Next we use `group_by` on `stations` using name and id, and summarize the following
(for both lat and lon):

- min 

- max

- range

- mean

```{r, cache = TRUE}
locations <- stations %>% 
  group_by(name, id) %>% 
  summarize(mean_lat = mean(lat),
            min_lat = min(lat),
            max_lat = max(lat),
            rng_lat = min_lat - max_lat,
            mean_lon = mean(lon),
            min_lon = min(lon),
            max_lon = max(lon),
            rng_lon = min_lon - max_lon)

locations %>% mutate_if(is.numeric, ~ round(.x, 4)) %>%  DT::datatable()
```

It turns out there are a few (11) records with `0` for lat or lon, we examine
these

```{r}
zero_stations <- stations %>% filter(lat == 0 | lon == 0) 
zero_stations
```

We remove records with 0 for lat or lon, then recalculate the mean lat and mean lon.
Additionally, we count the total number of trip starts and trip ends.  

```{r, cache = TRUE}
locations_summary <- stations %>% 
  filter(lat != 0, lon != 0) %>% 
  group_by(name) %>% 
  summarize(n = n(), 
            n_start = sum(type == "start"),
            n_end = sum(type == "end"),
            mean_lat = mean(lat),
            mean_lon = mean(lon))

glimpse(locations_summary)
```

Finally we write the `locations_summary` data frame to a .csv file to save it 
for future use.

```{r, cache = TRUE}
locations_summary %>% write_csv(here("results", "locations_summary.csv"))
```