---
title: "Mapping Ridership Data"
author: "Doug Antibus"
date: "2023-12-28"
output: 
  html_document:
    toc: true
    number_sections: false
    theme:
      version: 5
      bootswatch: pulse
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(cache = TRUE)
```


```{r}
source("00_libraries.r")
library(htmlwidgets)
library(leaflet)
library(htmltools)
```

### Introduction

Our random forest model indicated that **starting and ending stations** are highly
important in distinguishing casual riders from members.

In light of this, it makes sense to ask which specific stations are favored by 
members and casual riders. We will do this by using the `leaflet` package to make
interactive maps of start and end stations.  For each station, we will examine the 
*total number of trips* and whether the station is *favored by members or casual riders*.

We need to keep in mind that the dataset contains more records of member trips than 
casual trips.  To understand the bias towards member or casual ridership at each station, 
we will want to normalize data to the overall ratio of member:casual trips in the dataset.


### Data preparation

```{r, include = FALSE, cache = TRUE}
bikes <- read_csv(here("data", "bikes_2.csv")) %>% na.omit()
# bikes <- read_csv("C:/bikes_2.csv") %>%  na.omit()
```

Examine the bikes data frame.

```{r, cache = TRUE}
glimpse(bikes)
```

Begin by counting trips for each station by member/casual rider type.  We count
start and end records separately, then row-bind them into a single table, using the 
`type` variable to track starts and ends.

```{r, cache = TRUE}
start_counts <- bikes %>% count(start_name, member_casual) %>% 
  arrange(desc(n)) %>% 
  mutate(type = "start") %>% 
  rename(name = start_name)

end_counts <- bikes %>% count(end_name, member_casual) %>% 
  arrange(desc(n)) %>% 
  mutate(type = "end") %>% 
  rename(name = end_name)

stations <- bind_rows(start_counts, end_counts)
```

Next we create a wide-form data frame where member trips and casual trips are represented in separate columns.

```{r, cache = TRUE}
stations_w <- stations %>% 
  pivot_wider(values_from = n, names_from = (member_casual))
glimpse(stations_w)
```
We will need to account for the fact that member trips outnumber casual trips.  To
do this, we start by calculating the ratio of member:casual trips in the entire dataset, 
which will later be used to normalize data.

```{r}
ratio_member_casual <- bikes %>% 
  count(member_casual) %>% 
  {.$n[2] / .$n[1]}
ratio_member_casual
```

Now we perform transformations to prepare the data frame for plotting:

- calculate total trips and filter for total trips > 2000 to reduce overplotting.

- calculate the raw ratio of casual to member trips, then normalize this to account for the 
greater number of member trips in the dataset

- create a variable indicating whether ridership is biased towards members or casual riders

- calculate the log of the `normalized_ratio`.  This puts the data on a scale centered at 0 and improves the properties of the data for plotting

- create additional variables to assist with labeling

```{r}
stations_w <- stations_w %>% 
  mutate(total = member + casual) %>%
  filter(total > 2000) %>% 
  mutate(ratio = casual / member,
         normalized_ratio = ratio * ratio_member_casual, 
         bias_direction = ifelse(normalized_ratio <= 1, "member", "casual"),
         lr = log(normalized_ratio),
         bias = ifelse(bias_direction == "member", 1 / normalized_ratio,  normalized_ratio),
         col = ifelse(bias_direction == "member", "80714e", "#4d9275"),
         big = format(total, big.mark = ","))
```

Next we read in location data (mean latitude and longitude) that we previously calculated 
for each station and join this to the trip count data.

```{r}
locations <- read_csv(here("results", "locations_summary.csv")) %>% 
  select(name, mean_lat, mean_lon)

stations_w <- stations_w %>% left_join(locations, by = "name")
stations_w %>% mutate_if(is.numeric, ~ round(.x, 4)) %>%  DT::datatable()

```

After filtering for stations with > 2000 records, we can see how many start and end 
records we have for plotting.

```{r, echo = FALSE}
tab <- stations_w %>% 
  count(type) %>%  
  arrange(desc(type)) %>% 
  setNames(c("Type", "Count"))

tab
```




### Mapping rider type bias by station

With the data transformed we can finally put it onto a map!  

We'll map trip starts and ends separately, beginning with starts.  For these maps we'll
use size to represent overall trip count, and color to represent the degree of 
casual:member ridership bias (log of the normalized ratio of casual:member trips).  A log ratio of 0 would represent a ratio of 1, i.e. an equal tendency towards ridership by members
and casual riders.  A positive log ratio indicates a tendency towards casual ridership
over member ridership.

There are clear differences in how members and casual riders use stations 
for starting and ending trips.  The stations most highly favored by casual riders are near 
the lakeshore and/or adjacent to parks or other tourist destinations.  Some of these
stations are among those with the highest number of total trips.  For example, 
Streeter Drive & Grand Ave (near the Navy Pier) shows a 4.9 fold bias towards casual 
ridership for trip starts and a 6.3 fold bias towards casual riders for trip ends,
and has the most trip starts of any station.

Stations showing a bias towards member trips are mostly downtown.

Overall, the maps are consistent with the hypothesis that casual riders largely 
represent tourists, while members include more commuters.  Commuters might include 
people commuting to and from work or students.  For example, stations near 
the Illinois Institute of Technology and University of Illinois 
display a bias towards member ridership.

```{r start_map, include = FALSE}
starts_w <- stations_w %>% filter(type == "start")
pal3 <- colorNumeric(
  palette = "BrBG",
  domain = -2:2,
  reverse = FALSE
)

pal3_rev <- colorNumeric(
  palette = "BrBG",
  domain = -2:2,
  reverse = TRUE
)


starts_map <- leaflet(options = leafletOptions(minZoom = 10, maxZoom = 17)) %>% 
  addTiles() %>%
  addCircleMarkers(
    data = starts_w,
    lng = ~ mean_lon,
    lat = ~ mean_lat, 
    radius = ~ total / 1000,
    fillColor = ~ pal3(lr),
    stroke = TRUE,
    color = "black",  # stroke color
    weight = 2, # stroke weight
    fillOpacity = 0.8, 
    popup = paste0("<b style = color:Tomato;>",starts_w$name, "</b>",  # set color of name
                    "<br> <b>", starts_w$big, "</b> total trips",
                    "<br><b>", round(starts_w$bias, 2), "</b>", "X bias towards", " <b style = color:", starts_w$col, ">", starts_w$bias_direction, "</b>", " ridership",
                    "<br><b>", round(starts_w$lr, 1), "</b> log ratio of casual:member ridership"))   %>% 
  addLegend("topright", 
            pal = pal3_rev, 
            values = c(-2:2), 
            bins = 5, 
            # labels = c(-2, -1, 0, 1, 2),
            labFormat = labelFormat(transform = function(x) -1 * x),
            # title = "Casual / member <br> ridership (log ratio)"
            title = "Casual:Member <br>bias (log ratio)<br>") %>% 
            # title = "Log of ratio <br>(casual / member) <br>ridership") %>% 
  addControl("<h5>Rider types use different stations for starting trips</h5>", position = "bottomleft")
starts_map

```

```{r}
starts_map
```



```{r end_map, include = FALSE}
ends_w <- stations_w %>% filter(type == "end")
pal3 <- colorNumeric(
  palette = "BrBG",
  domain = -2:2,
  reverse = FALSE
)

pal3_rev <- colorNumeric(
  palette = "BrBG",
  domain = -2:2,
  reverse = TRUE
)

ends_map <- leaflet(options = leafletOptions(minZoom = 10, maxZoom = 17)) %>% 
  addTiles() %>%
  addCircleMarkers(
    data = ends_w,
    lng = ~ mean_lon,
    lat = ~ mean_lat, 
    radius = ~ total / 1000,
    fillColor = ~ pal3(lr),
    stroke = TRUE,
    color = "black",  # stroke color
    weight = 2, # stroke weight
    fillOpacity = 0.8, 
    popup = paste0("<b style = color:Tomato;>",ends_w$name, "</b>",  # set color of name
                   "<br> <b>", ends_w$big, "</b> total trips",
                   "<br><b>", round(ends_w$bias, 2), "</b>", "X bias towards", " <b style = color:", ends_w$col, ">", ends_w$bias_direction, "</b>", " ridership",
                   "<br><b>", round(ends_w$lr, 1), "</b> log ratio of casual:member ridership"))   %>% 
  addLegend("topright", 
            pal = pal3_rev, 
            values = c(-2:2), 
            bins = 5, 
            # labels = c(-2, -1, 0, 1, 2),
            labFormat = labelFormat(transform = function(x) -1 * x),
            # title = "Casual / member <br> ridership (log ratio)"
            title = "Casual:Member <br>bias (log ratio)<br>") %>% 
  # title = "Log of ratio <br>(casual / member) <br>ridership") %>% 
  addControl("<h5>Rider types use different stations for ending trips</h5>", position = "bottomleft")
ends_map
```

```{r}
ends_map
```


